
---
title: "Outlier Detection Techniques"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document demonstrates various advanced techniques for detecting outliers in a dataset. These include statistical methods (Z-Score, Elliptic Envelope), distance-based methods (KNN, Isolation Forest), and density-based methods (DBSCAN, LOF).

---

## Step 1: Load the Data

```{r load-data}
# Load necessary libraries
library(tidyverse)
library(FNN)
library(mvtnorm)
library(dbscan)
library(DMwR)
library(robustbase)
library(isolationForest)

# Load the dataset (assuming it's in the same folder as this Rmd file)
data <- read.csv("outlier_detection_data.csv")

# Quick look at the data
head(data)
summary(data)
```

---

## Step 2: Statistical Methods

### Z-Score
```{r z-score}
# Compute Z-Scores
data$z_score_feature_1 <- abs(scale(data$Feature_1))
data$z_score_feature_2 <- abs(scale(data$Feature_2))

# Flag outliers based on Z-Score threshold (e.g., > 3)
data$z_outliers <- ifelse(data$z_score_feature_1 > 3 | data$z_score_feature_2 > 3, TRUE, FALSE)
table(data$z_outliers)
```

### Elliptic Envelope
```{r elliptic-envelope}
# Fit the Elliptic Envelope model
env_model <- covMcd(data[, c("Feature_1", "Feature_2")])
env_outliers <- mahalanobis(data[, c("Feature_1", "Feature_2")], env_model$center, env_model$cov) > qchisq(0.975, 2)
table(env_outliers)
```

---

## Step 3: Distance-Based Methods

### K-Nearest Neighbors (KNN)
```{r knn}
# Compute distances to k nearest neighbors
knn_distances <- knnx.dist(data[, c("Feature_1", "Feature_2")], k = 5)
data$knn_outliers <- apply(knn_distances, 1, function(x) mean(x) > quantile(knn_distances, 0.95))
table(data$knn_outliers)
```

### Isolation Forest
```{r isolation-forest}
# Fit Isolation Forest
iso_model <- isolationForest$new()
iso_model$fit(data[, c("Feature_1", "Feature_2")])
data$isolation_outliers <- iso_model$predict(data[, c("Feature_1", "Feature_2")])
table(data$isolation_outliers)
```

---

## Step 4: Density-Based Methods

### DBSCAN
```{r dbscan}
# Fit DBSCAN
db <- dbscan::dbscan(data[, c("Feature_1", "Feature_2")], eps = 5, minPts = 5)
data$dbscan_outliers <- db$cluster == 0
table(data$dbscan_outliers)
```

### Local Outlier Factor (LOF)
```{r lof}
# Compute LOF
lof_scores <- lof(data[, c("Feature_1", "Feature_2")], k = 5)
data$lof_outliers <- lof_scores > quantile(lof_scores, 0.95)
table(data$lof_outliers)
```

---

## Conclusion

This document demonstrates multiple approaches to outlier detection, highlighting their strengths and limitations. Choose the most appropriate method based on your dataset and analysis goals.

